A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets

### 2012-10-26 manuscript/cloudbreak.tex
Add new ref

Genomic structural variation

Difficulty of detection.. Lack of concordance of current algorithms...

Typical algorithms use only unambiguous discordant read pair mappings...

Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \cite{Schatz:2009p278} querying variant databases, \cite{Oconnor:2010p1835} single nucleotide polymorphism (SNP) calling, \cite{Langmead:2009p1225} RNA-seq differential expression analysis, \cite{Langmead:2010p1268} ChIP-seq peak calling, \cite{Feng:2011p1228} and computing genome mappability. \cite{Lee:2012bk} The MapReduce programming style lends itself particularly well towards resequencing tasks such as SNP calling, in which a result can be computed independently for various locations along the genome. As mentioned previously, most algorithms for SV calling work by clustering discordant read pairs across the genome, an operation that does not lend itself to easy parallelization. Cloudbreak instead independently computes local scores for small contiguous regions along the genome. The recently published method forestSV \cite{Michaelson:2012fj} showed that SV detection from computation of local scores is a promising approach, since multiple features can be extracted from reads aligning to or near each region, and integrated with features built from genome annotations in machine learning frameworks. This approach also allows the use of a MapReduce programming style and opens the door to cluster computing for SV detection.

\section{Methods}\label{Methods}

\subsection{A Hadoop/MapReduce Algorithm for SV Detection}

@@

Cloudbreak divides its processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes local SV scores for a set of small windows completely covering the target genome. Finally, the SV scores are exported from the Hadoop file system and post-processed to produce a final set of SV calls which span multiple genomic windows. The Cloudbreak MapReduce algorithm is summarized in \ref{algo1}.

@@

The Cloudbreak algorithm could use any aligner that is capable of reporting multiple alignments for reads. We report results generated with alignments from Novoalign. \cite{novoalign} Novoalign exhibits higher sensitivity for mapping locations than most aligners, even in the presence of indels and errors, at the expense of greater runtime. \cite{Krawitz:2010iq} In addition, Novoalign can report multiple possible mapping locations for a read and has accurate mapping quality scores \cite{Ruffalo:2011p1758}, which are useful in assigning confidence scores to SV calls. The current implementation is also capable of using MrFast \cite{Alkan:2009cr} as an aligner, although MrFast does not provide mapping quality scores, which makes Cloudbreak predictions less accurate.

In the second job, mappers collect information, namely the insert size and the pair mapping quality, about all of the possible alignments of read pairs that span each genomic locus, defined as a small (25bp) window in the genome reference. Reducers then integrate all of the read pair information for reads that span a given location to produce an SV score indicating the likelihood of an SV affecting that locus. Since scores are computed independently at each locus, this procedure is essentially constructing and analyzing a pileup of reads pairs at each location, making it similar to previous bioinformatic MapReduce algorithms for SNP calling implemented in Crossbow \cite{Langmead:2009p1225} and GATK. \cite{McKenna:2010p1051}

@@

As described in the previous section, reducers in the second Cloudbreak job compute an SV score indicating the presence or absence of a deletion for each genomic location $l$. Locations are defined as contiguous windows of 25bp covering the entire genome. For each location, Cloudbreak examines the set of possible alignments of read pairs that spans that location, meaning that either of the reads or the insert between them overlaps the location. We begin with the prior probabilities of having or not having a deletion that spans the location, $P(D)$ and $P(\bar{D}) = 1 - P(D)$. Based on the results reported by the 1000 Genomes Project pilot data \cite{Mills:2011p1611} we estimate $P(D)$ as approximately $9x10^{-7}$. Each possible spanning alignment indicates an implied insert size $s$ based on the distance between the alignments of the two ends, as well as a mapping quality that indicates $P(m)$, or the probability that the pair alignment is correct. Based on $s$, for each possible pair alignment we compute $P(s|D)$ and $P(s|\bar{D})$, or the probability of observing an aligning pair with insert size $s$ in the presence or absence of a deletion in the current window. $P(s|{\bar{D}})$ can be estimated using a normal distribution with a mean at the expected fragment size of the read group to which the current pair belongs. We estimate $P(s|D)$ using a lognormal distribution fit to the published range of deletion sizes from the 1000 Genomes Project pilot data. Using our estimated prior probabilities we then compute the quantities

@@

\section{Results}\label{results}

@@

There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J. Craig Venter \cite{Levy:2007fb} and applied them to the human GRCh36 chromosome 2 and 17 reference sequences. We then simulated 50bp paired Illumina reads with a fragment size of 200bp from this modified sequence using \emph{dwgsim} from the DNAA software package \cite{DNAA} at varying levels of coverage, and compared results from Cloudbreak with the original list of deletions. A similar approach was recently used by \cite{Sindi:2012kk} to evaluate the performance of GASVPro.

@@

>>>>

Genomic structural variation.. as much or more variance as SNPS.. INDELS in exons, coding regions (Mills 2011)...

Difficulty of detection.. Lack of concordance of current algorithms (described in 1KG SV paper)...

Typical algorithms use only unambiguous discordant read pair mappings... New algos combine RP and RD or SP signals...

Use of the Hadoop framework has been demonstrated for sequencing-related bioinformatic tasks including short read mapping, \cite{Schatz:2009p278} querying variant databases, \cite{Oconnor:2010p1835} single nucleotide polymorphism (SNP) calling, \cite{Langmead:2009p1225} RNA-seq differential expression analysis, \cite{Langmead:2010p1268} ChIP-seq peak calling, \cite{Feng:2011p1228} and computing genome mappability. \cite{Lee:2012bk} The MapReduce programming style lends itself particularly well towards resequencing tasks such as SNP calling, in which a result can be computed independently for various locations along the genome. As mentioned previously, most algorithms for SV and INDEL calling work by clustering discordant read pairs across the genome, an operation that does not lend itself to easy parallelization. Cloudbreak instead independently computes local scores for small contiguous regions along the genome. The recently published method forestSV \cite{Michaelson:2012fj} showed that SV detection from computation of local scores is a promising approach, since multiple features can be extracted from reads aligning to or near each region, and integrated with features built from genome annotations in machine learning frameworks. This approach also allows the use of a MapReduce programming style and opens the door to cluster computing for SV detection.

\section{Results}\label{Results}

\subsection{A Hadoop/MapReduce Algorithm for INDEL Detection}

@@

Cloudbreak divides its processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes local deletion scores for a set of small windows completely covering the target genome. Finally, the scores are exported from the Hadoop file system and post-processed to produce a final set of variant calls which span multiple genomic windows. The Cloudbreak MapReduce algorithm is summarized in \ref{algo1}.

@@

The Cloudbreak algorithm could use any aligner that is capable of reporting multiple alignments for reads. We report results generated with alignments from Novoalign. \cite{novoalign} Novoalign exhibits higher sensitivity for mapping locations than most aligners, even in the presence of indels and errors, at the expense of greater runtime. \cite{Krawitz:2010iq} In addition, Novoalign can report multiple possible mapping locations for a read and has accurate mapping quality scores \cite{Ruffalo:2011p1758}, which are useful in assigning confidence scores to variant calls. The current implementation is also capable of using MrFast \cite{Alkan:2009cr} as an aligner, although MrFast does not provide mapping quality scores, which makes Cloudbreak predictions less accurate.

In the second job, mappers collect information, namely the insert size and the pair mapping quality, about all of the possible alignments of read pairs that span each genomic locus, defined as a small (25bp) window in the genome reference. Reducers then integrate all of the read pair information for reads that span a given location to produce a score indicating the likelihood of a deletion affecting that locus. Since scores are computed independently at each locus, this procedure is essentially constructing and analyzing a pileup of reads pairs at each location, making it similar to previous bioinformatic MapReduce algorithms for SNP calling implemented in Crossbow \cite{Langmead:2009p1225} and GATK. \cite{McKenna:2010p1051}

@@

As described in the previous section, reducers in the second Cloudbreak job compute a score indicating the presence or absence of a deletion for each genomic location $l$. Locations are defined as contiguous windows of 25bp covering the entire genome. For each location, Cloudbreak examines the set of possible alignments of read pairs that spans that location, meaning that either of the reads or the insert between them overlaps the location. We begin with the prior probabilities of having or not having a deletion that spans the location, $P(D)$ and $P(\bar{D}) = 1 - P(D)$. Based on the results reported by the 1000 Genomes Project pilot data \cite{Mills:2011p1611} we estimate $P(D)$ as approximately $9x10^{-7}$. Each possible spanning alignment indicates an implied insert size $s$ based on the distance between the alignments of the two ends, as well as a mapping quality that indicates $P(m)$, or the probability that the pair alignment is correct. Based on $s$, for each possible pair alignment we compute $P(s|D)$ and $P(s|\bar{D})$, or the probability of observing an aligning pair with insert size $s$ in the presence or absence of a deletion in the current window. $P(s|{\bar{D}})$ can be estimated using a normal distribution with a mean at the expected fragment size of the read group to which the current pair belongs. We estimate $P(s|D)$ using a lognormal distribution fit to the published range of deletion sizes from the 1000 Genomes Project pilot data. Using our estimated prior probabilities we then compute the quantities

@@

@@

There is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J. Craig Venter \cite{Levy:2007fb} and applied them to the human GRCh36 chromosome 2 and 17 reference sequences. We then simulated 50bp paired Illumina reads with a fragment size of 200bp from this modified sequence using \emph{dwgsim} from the DNAA software package \cite{DNAA} at varying levels of coverage, and compared results from Cloudbreak with the original list of deletions. A similar approach was recently used by \cite{Sindi:2012kk} to evaluate the performance of GASVPro.

@@

\section{Methods}\label{methods}
